---
title: "Borrow predictions, then calibrate in the trial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Borrow predictions, then calibrate in the trial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(roscar)
```

## A complete worked pass

The external data provide an outcome prediction. Randomized treatment and trial
outcomes still supply the treatment-effect estimate. This example follows
Algorithm 1 using four independent standard-normal covariates and known treatment
probability one half. The trial outcome is

\[
Y=1+X_1+X_2+A(0.5+0.5X_1)/2+\varepsilon,
\]

where treatment is coded as minus or plus one and the error is standard normal.
The independent external outcome follows the same model with a level shift of
0.5. The analytic conditional treatment effect is \(0.5+0.5X_1\).

```{r example-data}
sim <- simulate_scenario("teaching", n_r = 200, n_o = 2000,
                         seed = 20260914)
learners <- list(external = learner_ols(), calibration = learner_intercept(),
                 trial = learner_ols(), final = learner_ols())
fit <- borrow_cate(sim$trial, sim$ext, recipe = "A", learners = learners,
                   basis = ~X1, K = 5, seed = 20260914)
```

The external regressions fit each treatment arm independently. Within each trial
training fold, intercept calibration estimates how far the external outcome
level is displaced. Its prediction is evaluated on the held-out fold. The final
regression uses an intercept and `X1`, the prespecified effect basis.

```{r intermediate-rows}
h <- head(fit$heldout, 6)
h$external_prediction <- ifelse(h$A == 1, h$q_plus, h$q_minus)
h[, c("fold", "A", "pi", "external_prediction", "q_plus", "q_minus",
      "mu_cal_plus", "mu_cal_minus", "baseline", "pseudo_outcome")]
```

The personalized baseline combines calibrated arm predictions with the
opposite-arm assignment probabilities. With equal allocation it is their
average. The pseudo-outcome is \(A(Y-\widehat m)/\pi\), where `pi` is the
probability of the treatment actually received. Multiplying by treatment gives
the sign; dividing by one half supplies the factor of two.

```{r predictions}
data.frame(X1 = c(-1, 0, 1), analytic_truth = sim$truth_profiles,
           recipe_A = predict(fit, sim$profiles),
           trial_only = predict(fit$trial_only, sim$profiles))
c(recipe_A = var(fit$heldout$pseudo_outcome),
  trial_only = var(fit$heldout$psi_trial))
```

## Conditional bootstrap intervals

The publication example reruns the complete trial pipeline 500 times, resampling
independent trial units and keeping the external data fixed. Repeated copies of
one sampled unit stay together in the cross-fitting folds. The same resamples
also refit the trial-only comparator. This code is left unevaluated during a
routine documentation build because the reproduction script saves the complete
500-resample result separately.

```{r bootstrap, eval=FALSE}
fit <- borrow_cate(sim$trial, sim$ext, recipe = "A", learners = learners,
                   basis = ~X1, K = 5, B = 500, grid = sim$profiles,
                   seed = 20260914)
fit$bootstrap$intervals
```

<!-- BEGIN VERIFIED TEACHING OUTPUT -->
The recorded 500-resample run gives the following pointwise 95% intervals.

| X1 | Analytic truth | Recipe A (95% interval) | Trial-only (95% interval) | Width ratio |
|---:|---:|:---|:---|---:|
| -1 | 0 | 0.195 (-0.212, 0.603) | 0.207 (-0.203, 0.608) | 1.005 |
| 0 | 0.5 | 0.724 (0.454, 1.019) | 0.752 (0.467, 1.044) | 0.978 |
| 1 | 1 | 1.252 (0.838, 1.622) | 1.296 (0.825, 1.697) | 0.898 |

The held-out pseudo-outcome variance is 4.508 for Recipe A and 4.591 for the paired trial-only fit. Width ratios above one indicate increased uncertainty; ratios below one indicate reduced uncertainty.

<!-- END VERIFIED TEACHING OUTPUT -->

From the repository root, `source("paper/teaching/run.R"); run_teaching()` writes
the call, the first six intermediate rows, all three profile estimates and
intervals, paired interval-width ratios, and the four pseudo-outcome variances
to `paper/teaching/output/`. Analytic truths are printed separately. A width ratio
above one reports an increase in uncertainty. One dataset does not establish
coverage or a typical gain from borrowing.

## External controls and bundled data

Recipe B borrows a single external arm. Its other outcome model is learned
entirely from the trial. The package also retains the original bundled datasets
and the `cate_model()` interface.

```{r bundled}
data("rct", "os", package = "roscar")
trial <- trial_data(rct$X, rct$A, rct$Y, pi = 0.5)
external <- external_data(os$X, os$A, os$Y)
fit_B <- borrow_cate(trial, external, recipe = "B", arm_ext = -1,
                     learners = learners, basis = ~X1, K = 5, seed = 42)
head(predict(fit_B, trial$X))
```

## Covariate mismatch and diagnostics

Recipe C predicts the external-only covariates from shared covariates, then
calibrates the transported predictions in the trial. Recipe D uses the linear
CALM outcome-supervised embedding. Its pooled external PLS encoder excludes
treatment. Use the same final effect basis for comparisons.

```{r mismatch, eval=FALSE}
mismatch <- simulate_scenario(4, rho = 0.8, seed = 42)
fit_C <- borrow_cate(mismatch$trial, mismatch$ext, recipe = "C",
                     basis = mismatch$basis, seed = 42)
fit_D <- borrow_cate(mismatch$trial, mismatch$ext, recipe = "D",
                     basis = mismatch$basis, seed = 42)
plot(diagnose(fit_C))
```

Inspect held-out arm losses and the borrowing diagnostic alongside uncertainty
and effect estimates. Better external prediction alone does not guarantee a
better conditional treatment-effect estimate.
